Model Finding for Finite Function Traces in SMT

### 2015-04-28 tex/main.tex
More work, add example.

\newcommand{\ssorts}[1]{#1^\mathrm{s}}
\newcommand{\sfuns}[1]{#1^\mathrm{f}}
\newcommand{\sfundefs}[1]{#1^\mathrm{df}}
\newcommand{\sfunndefs}[1]{#1^\mathrm{nf}}

@@

@@

\title{Finite Traces of Recursive Functions in SMT}

@@

Leon, others?

@@

A \emph{signature} $\Sigma$ consists of 
a set $\ssorts{\Sigma}$ of sort symbols and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean sort $\Bool$ and constants 
$\ltrue$ and $\lfalse$ of that sort (respectively, for true and false), 
an (infix) equality predicate $\teq$ of type $\sigma \times \sigma$ for each $\sigma \in \ssorts{\Sigma}$,
standard boolean connectives ($\wedge$, $\vee$, $\Leftrightarrow$, etc.),
and a predicate $\lite$ of type $\Bool \times \sigma \times \sigma$ for each $\sigma \in \ssorts{\Sigma}$.

@@

\begin{definition}
A formula $\varphi$ is well-defined with respect to functions $S$ if it is of the form:
\begin{equation*}
(\forallf{f_1} \vec x_1. \varphi_1) \wedge \ldots \wedge (\forallf{f_n} \vec x_n. \varphi_n) \wedge \psi
\end{equation*}
where $f_1, \ldots, f_n$ are distinct, 
$S = \{ f_1, \ldots, f_n \}$,
and $\psi$ contains no function definitions.
\end{definition}

Given a signature $\Sigma$, 
we parition its function symbols $\sfuns{\Sigma}$ into $\sfundefs{\Sigma} \mathrel{\uplus} \sfunndefs{\Sigma}$,
where $\sfundefs{\Sigma}$ is the set of \emph{defined} function symbols,
and $\sfunndefs{\Sigma}$ are the others.
In this paper, we will consider $\Sigma$-formulas that are well-defined for $\sfundefs{\Sigma}$.

We write $\terms( t )$ to denote the set of subterms in $t$.
We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$, where $f \in \sfundefs{\Sigma}$.

\subsection{Encoding Well-Defined Assumption for Recursive Functions}

In the remainder of the section, we fix a signature $\Sigma$.

The main idea behind this translation is to abstract the set of tuples that a recursive function $f$ is invoked on as a (fresh) uninterpreted sort $\fargsort{f}$.
We extend our signature $\Sigma$ as follows.
For each $f \in\sfundefs{\Sigma}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$, we associate:
\begin{enumerate}
\item[-] an uninterpreted sort $\fargsort{f}$, and
\item[-] a vector of $n$ uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ of type $(\fargsort{f} \rightarrow \tau_1)$, $\ldots$, $(\fargsort{f} \rightarrow \tau_n)$.
\end{enumerate}

The role of the uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ will be discussed more in the following.

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( t^\tau, p )$ : 
 \begin{itemize}
  \item[] $\mathsf{if}$ $t \equiv \forallf{f} \vec x. t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv( t_1, p )$
      \item[] $\mathsf{if}$ $f \equiv \fnull$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x. r_1, \forall \vec x. D_1 )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        \item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y^{\fargsort{f}}. ( r_1 [ \vecfarg{f}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z^{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \functerms^\Sigma( t ) \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is well-defined with respect to function definitions.
In the procedure, $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$.
}
\label{fig:encoding}
\end{figure}

Figure~\ref{fig:encoding} gives a procedure $\conv$ for translating a formula $t$ containing function definitions.
The procedure takes two arguments: the term $t^\tau$ to translate, and a \emph{polarity} p (either $\ptrue$, $\pfalse$, or $\pnone$).
If function definition quantifiers $\forallf{f}$ in $t$ are such that $f$ is well-defined and $\conv( t, \ptrue )$ returns $( r, \emptyset )$,
then $r$ is equivalent to $t$.

\begin{example}
Consider the formula
\end{example}

>>>>

\newcommand{\ssorts}[1]{#1^\mathrm{s}}
\newcommand{\sfuns}[1]{#1^\mathrm{f}}
\newcommand{\sfundefs}[1]{#1^\mathrm{df}}
\newcommand{\sfunndefs}[1]{#1^\mathrm{nf}}
\newcommand{\sortint}{\mathrm{Int}}

@@

\newcommand{\Sigmalia}{\Sigma_{LIA}}
\newcommand{\extendsig}[1]{\mathcal{E}( #1 )}

\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
\newcommand{\ajr}[1]{\rem{#1 --ajr}}

\newcommand{\vthinspace}{\kern+0.083333em}
\newcommand{\typ}[1]{^{\vthinspace #1}}

@@

\title{Model Finding for Finite Function Traces in SMT}

@@

Leon, Nitpick, others?

@@

A \emph{signature} $\Sigma$ consists of 
a set $\ssorts{\Sigma}$ of sort symbols and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean sort $\Bool$ and constants 
$\ltrue$ and $\lfalse$ of that sort (respectively, for true and false), 
an (infix) equality predicate $\teq$ of type $\sigma \times \sigma$ for each $\sigma \in \ssorts{\Sigma}$,
and standard boolean connectives ($\neg$, $\wedge$, $\vee$, $\lite$, etc.).
%and a function $\lite$ of type $\Bool \times \sigma \times \sigma$ for each $\sigma \in \ssorts{\Sigma}$.
We will write $t\typ{\tau}$ to denote a term $t$ of type~$\tau$.

@@

\ajr{theories, models, etc.}

\begin{definition}
A formula $\varphi$ \emph{well-defined with respect to functions} $S$ if it is of the form:
\begin{equation*}
(\forallf{f_1} \vec x_1. \varphi_1) \wedge \ldots \wedge (\forallf{f_n} \vec x_n. \varphi_n) \wedge \psi
\end{equation*}
where $f_1, \ldots, f_n$ are distinct, 
$S = \{ f_1, \ldots, f_n \}$,
and $\psi$ contains no function definitions.
\end{definition}

Given a signature $\Sigma$, 
we will commonly parition its function symbols $\sfuns{\Sigma}$ into $\sfundefs{\Sigma} \mathrel{\uplus} \sfunndefs{\Sigma}$,
where $\sfundefs{\Sigma}$ is the set of \emph{defined} function symbols,
and $\sfunndefs{\Sigma}$ are the others,
and consider $\Sigma$-formulas that are well-defined with respect to functions $\sfundefs{\Sigma}$.

We write $\terms( t )$ to denote the set of subterms in $t$.
We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

\subsection{Encoding Well-Defined Assumption for Recursive Functions}
$\Leftrightarrow$, 
In the remainder of the section, we fix a signature $\Sigma$.

Say that $\Sigma$-formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma}$,
and the definitions in $\sfundefs{\Sigma}$ are logically consistent \ajr{formalize}.
In this section, we present a method which constructs an extended signature $\Sigma'$,
and translates $\varphi$ to equisatisfiable $\Sigma'$-formula $\psi$ on which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( t\typ{\tau}, p )$ : 
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t \equiv \forallf{f} \vec x. t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv( t_1, p )$
      \item[] $\mathsf{if}$ $f \equiv \fnull$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x. r_1, \forall \vec x. D_1 )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        \item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y\typ{\fargsort{f}}. ( r_1 [ \vecfarg{f}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \functerms^\Sigma( t ) \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is well-defined with respect to function definitions.
In the procedure, $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$,
where $\boolop$ is an (interpreted) predicate.
}
\label{fig:encoding}
\end{figure}

The intuition behind this translation is use an uninterpreted sort $\fargsort{f}$ to abstract the set of tuples that each recursive function $f$ is invoked on,
and restrict the quantification of the function definition for $f$ to a variable of sort $\fargsort{f}$.
In more detail, we construct an extended signature of $\Sigma$, which we will write as $\extendsig{\Sigma}$, as follows.
For each $f \in \sfundefs{\Sigma}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$, 
the signature $\extendsig{\Sigma}$ includes:
\begin{enumerate}
\item[-] an uninterpreted sort $\fargsort{f}$, and
\item[-] a vector of $n$ uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ of type $(\fargsort{f} \rightarrow \tau_1)$, $\ldots$, $(\fargsort{f} \rightarrow \tau_n)$.
\end{enumerate}
%The interpretation of uninterpreted sort $\fargsort{f}$ will denote the elements (tuples) on which the function $f$ is applied.
%The role of the uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ will be discussed more in the following.
We then run the procedure $\conv$ in Figure~\ref{fig:encoding} on $\varphi$.
The procedure $\conv$ takes two arguments: the term $t$ to translate, and a \emph{polarity} $p$ (either $\ptrue$, $\pfalse$, or $\pnone$).
It returns a pair of the form $( r, D )$, where $r$ is a term of type $\tau$, and $D$ is a set of formulas, which we will call its \emph{domain constaints}.
The role of $D$ is to ensure that the (restricted) function definition for $f$ includes certain tuples in its domain, which we explain more in the following.

In the case that $t$ is an application of (interpreted) predicate $\boolop$ (e.g. a boolean connective, or equality),
we recursively call $\conv$ on its subchildren $t_i$ and polarity $\pol( \boolop, i, p )$, where $\pol$ is defined as\footnote{The negation $\neg p$ of polarity $p$ is 
$\pfalse$ if $p$ is $\ptrue$,
$\ptrue$ if $p$ is $\pfalse$, and
$\pnone$ if $p$ is $\pnone$.}:
\begin{equation*}
\pol( \boolop, i, p ) = \begin{cases}
                         \pnone & \boolop \equiv \teq \text{ or } \boolop \equiv \lite \wedge i=0 \\
                         \neg p & \boolop \equiv \neg \\
                         p & \text{otherwise}
                         \end{cases}
\end{equation*}
If $t$ has positive polarity,
we construct a conjunction of the result of the recursive calls with domain constraints $D_1, \ldots, D_n$.
Similarly, if $t$ has negative polarity,
we construct a disjunction of the result of the recursive calls with the negation of domain constraints $D_1, \ldots, D_n$.
It $t$ has no polarity,
then we return the union of the domain constraints $D_1 \cup \ldots \cup D_n$.

In the case that $t$ is a quantified formula $\forallf{f} \vec x. t_1$, it recursively calls $\conv$ on its body with the same polarity.
In the case that $t$ is not a function definition, 
the resulting quantified formula is reconstructed with the term $r_1$ returned by the recursive call,
and a quantified prefix is appended to the domain constraints $D_1$ returned by the recursive call.
In the case that $t$ is a function definition,
we instead construct a quantified formula over a single variable $y$ of type $\fargsort{f}$,
and replace all occurrences of $\vec x$ in $r_1$ with $\vecfarg{f}( y )$.
In this case, since function definitions are a top-level conjunct, by case analysis on the return values of $\conv$, we know that $D_1$ is empty.

We demonstrate this translation with a simple example.

\begin{example}
Consider the (combined) signature $\Sigma = \Sigmalia \cup \Sigma_u$,
where $\ssorts{\Sigma_u} = \{ \sortint \}$, and 
$\sfuns{\Sigma_u} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$,
and the $\Sigma$-formula $\varphi$:
\begin{equation*}
\forall_{s} x\typ{\sortint}. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{equation*}
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
The signature $\extendsig{\Sigma}$ includes the sort $\fargsort{s}$,
and the uninterpreted function $\farg{s}\typ{ \fargsort{s} \rightarrow \sortint }$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{equation*}
\begin{split}
\forall y\typ{\fargsort{s}}. ite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( a ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq a )
\end{equation*}
Note that the universal quantification in this example (the left disjunct) is $\ldots$
$\square$
\end{example}

We claim that if $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma}$,
function definitions in $\sfundefs{\Sigma}$ are logically consistent \ajr{formalize},
and $\conv( \varphi, \ptrue )$ returns $( \psi, \emptyset )$,
then $\psi$ is equisatisfiable to $\varphi$.

\begin{lemma}
Let $\varphi$ be a $\Sigma$-formula that is well-defined with respect to functions $\sfundefs{\Sigma}$,
and let $( \psi, D )$ be the result of $\conv( \varphi, \ptrue )$.
Then,
(i) $D = \emptyset$, and
(ii) for each model of $\varphi$, $\ldots$. \ajr{todo}
\end{lemma}
